---
title: "Investigating whether closer races lead to higher turnouts"
author: "Chengyang Li"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---


# Description

In trying to understand voting behaviour, it has been conjectured that closer races lead to higher turnouts. We investigate this conjecture with the 2018 Ottawa municipal election data.

# The closer races lead to higher turnouts
```{r, warning=FALSE, message=FALSE}
# load libraries
library(dplyr)
library(readxl)
library(janitor)
library(ggplot2)
library(knitr)
library(ggrepel)
library(stringr)
library(tidyr)
library(readr)

# getting data of turnouts

df <- read_excel("statementofvotescastoctober242018.xls", sheet = 1, skip=4) %>%
  clean_names() %>% 
  dplyr::select(-starts_with("x")) %>%
  filter(!is.na(registered_voters)) %>%
  filter( precinct != "City / Ville - Total") %>%
  filter(!str_detect(precinct, '^Spc Adv')) %>%
  filter(!str_detect(precinct, '^Adv')) %>%
  mutate( ward = str_sub(precinct,1,2)) %>%
  group_by(ward) %>%
  summarize(
    total_votes = sum(cards_cast),
    total_voters = sum(registered_voters)
  ) %>%
  mutate( fraction_voted = total_votes/total_voters)
  
df <- df %>% select(ward, fraction_voted)

# getting the data of voting difference between  winner and the 2nd winner

theWin <- function(ward){
  df2 <- read_excel("statementofvotescastoctober242018.xls", sheet=ward+2, skip = 3) %>%
  clean_names() %>% 
  select(-starts_with("x"), -precinct_7) %>%
  filter(!is.na(times_cast)) %>%
  rename(precinct = precinct_1) %>%
  na.omit() 
  
  winner <- df2 %>% 
    select(-times_cast, -undervotes, -overvotes, -total_votes) %>%
    pivot_longer(-precinct, names_to = "candidate", values_to = "votes") %>%
    group_by(candidate) %>%
    summarize( total_votes = sum(votes)/2 ) %>%
    arrange(desc(total_votes)) %>%
    head(2) %>%
    select(total_votes) 
  inter <- (winner$total_votes[1] - winner$total_votes[2])

  return(tibble( inter = inter ))
}

# adding str_pad of the wards to the turnouts
dfwards <- tibble()
for (ward in 1:23) {
  ans <- theWin(ward)
  ans <- ans %>% mutate( 
    ward = str_pad(ward, width = 2, side='left', pad = '0') 
  )
  dfwards <- bind_rows(dfwards, ans)
}

# merge two dataframes
df_full <- dfwards %>% left_join(df, by = c("ward" = "ward"))

# make plot
plot <- df_full %>%
  ggplot(aes( x = inter, y = fraction_voted))+
  geom_point()+
  geom_smooth(method='lm', se=FALSE)+
  labs(
    x = 'Voting difference between 1st and 2nd in one ward',
    y = 'Turnout',
    title = 'The closer races lead to higher turnouts'
  )

plot

```

Another useful plot with label of the wards
```{r, warning=FALSE, message=FALSE}
plot_t <- df_full %>%
  ggplot(aes( x = inter, y = fraction_voted, label = ward))+
  geom_text_repel()+
  geom_point()+
  geom_smooth(method='lm', se=FALSE)+
  labs(
    x = 'Voting difference between 1st and 2nd in one ward',
    y = 'Turnout',
    title = 'The closer races lead to higher turnouts'
  )

plot_t
```

Another useful plot with relative size of turnouts
```{r, warning=FALSE, message=FALSE}
plot_s <- df_full %>%
  ggplot(aes( x = inter, y = fraction_voted, size = fraction_voted))+
  geom_point()+
  geom_smooth(method='lm', se=FALSE)+
  labs(
    x = 'Voting difference between 1st and 2nd in one ward',
    y = 'Turnout',
    title = 'The closer races lead to higher turnouts'
  )

plot_s

```


In this investigation from 2018 Ottawa municipal election data, the closer races lead to higher turnouts among 23 wards. A majority of wards which have closer races (voting difference between 1st and 2nd are less than 2500) lead to higher turnouts (turnouts between 35% and 40%). The rest of wards have over 2500 voting difference and below 32.5% turnouts.


# Comments

 * It is worth noting that there are two wards (15 and 03) does not have close race but leading to high turnouts.
 * The highest turnout almost leads the cloest race in one of 23 wards.
 * It is ad-hoc that make definition of race as winner's votes minus 2nd winner's votes.
 * No statistical tests were performed, hence any conclusions are merely speculation.
 
# Suggestions for further explorations

 * Perhaps there are more interesting details if we apply type of areas (urban, suburban and rural) to see exact relationship may differ by type of area.
 * Would be interesting to further investigate two wards that have higher turnouts but no closer races.
 
# References

 * R for Data Science

# Appendix: sessionInfo
```{r}
sessionInfo()
```